On probabilistic one-vs-one classification

### 2025-04-21 paper.tex
Fix definition of Bayes covariance

Suppose we have $K$ classes of objects, each distributed according to a probability distribution $p_i$, $i=1,\ldots, K$ on a feature space $X$. In a multi-class classification problem we are provided with the prior probabilities $\Pi = (P_1, \ldots, P_k)$ of each class. Given $x$ in $X$ we aim to find the probability that $x$ belongs to $i$-th class.

@@

\begin{assumption} \label{ass:1}
 For all $x$ in the feature space we have $p_i(x) > 0$.
\end{assumption}
Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines all satisfy this assumption.

The Bayes classifier predicts that the probability of $x$ belonging to the $i$-th class is
\begin{align}
 p^\Pi_i(x) = \frac{p_i(x) P_i}{\sum_{k=1}^K p_k(x)P_k}.
\end{align}

@@

\begin{align}
	p_{ij}^\Pi(x) = \frac{p_i(x) P_i}{p_i(x)P_i + p_j(x)P_j}.
\end{align}

@@

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow{dr}[swap]{multi-class~ prediction} & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Bayes covariance} is a notion introduced in  [\cite{vsuch2016bayes}]. It refers to the behavior of a probabilistic multi-class method built in one-vs-one fashion, when the priors change. Any Bayes classifier, whether binary or multiclass changes its prediction when the priors change. Let us describe the process. Suppose we change priors from $\Pi = (P_1, \ldots, P_k)$ to $\Pi'= (P'_1, \ldots, P'_K)$. Then the prediction of Bayes classifiers changes from $\boldsymbol{p}^\Pi= (p_1, \ldots, p_K)$ to the vector 
\begin{align}
\boldsymbol{p}^{\Pi'} \propto (p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}). \label{eq:changePrior}
\end{align}

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $\boldsymbol{R}$, then we say that the coupling method is Bayes covariant. Symbolically, let us denote by  $\pi$ the change of priors. Let $\boldsymbol{R}^\pi$ be pairwise datum when we apply \eqref{eq:changePrior} to each binary classifier.
For Bayes covariant method $\boldsymbol{v}$ the following diagram commutes.

\begin{figure}[!h]
\centering
\begin{tikzcd}
	\boldsymbol{R} \arrow[r, "\pi"] \arrow[d, "coupling"] & \boldsymbol{R}^\pi \arrow[d, "coupling"] \\
	\boldsymbol{v}(\boldsymbol{R}) \arrow[r, "\pi"] & \boldsymbol{v}(\boldsymbol{R}^\pi) =\boldsymbol{v}(\boldsymbol{R})^\pi
\end{tikzcd}
\caption{Diagrammatical description of Bayes covariance for a coupling method $\boldsymbol{v}$}
\label{fig:bc}
\end{figure}

@@

\begin{prop}
Any arboreal method is Bayes covariant.
\end{prop}

@@

Suppose real numbers $\alpha_1, \alpha_L$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are coupling methods. We define their linear combination $\bigoplus_i a_i \boldsymbol{v}_i$ as the coupling method that for any $x$ yields the probability distribution on $L$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_L \boldsymbol{v}_L\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_L^{a_L}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication. With this definition we have the following result.

@@

>>>>

Suppose we have $K$ classes of objects, each distributed with density $d_i$, $i=1,\ldots, K$ on a feature space $X$. In a multi-class classification problem we are provided with the prior probabilities $\Pi = (P_1, \ldots, P_k)$ of each class. Given $x$ in $X$ we aim to find the probability that $x$ belongs to $i$-th class.

@@

\begin{assumption} \label{ass:1}
 For all $x$ in the feature space we have $d_i(x) > 0$.
\end{assumption}
Commonly used classifiers such as linear discriminant analysis, (penalized) logistic regression or support vector machines all satisfy this assumption.

The Bayes classifier predicts that the probability of $x$ belonging to the $i$-th class is
\begin{align}
 p^\Pi_i(x) = \frac{d_i(x) P_i}{\sum_{k=1}^K d_k(x)P_k}.
\end{align}

@@

\begin{align}
	p_{ij}^\Pi(x) = \frac{d_i(x) P_i}{d_i(x)P_i + d_j(x)P_j}.
\end{align}

@@

\begin{figure}[!h]
	\centering
	\begin{tikzcd}[column sep=3cm]
		\textrm{sample $x$}\arrow[r, "binary~classifiers"] \arrow[dr,dotted,swap, "multi-class~ prediction"] & \boldsymbol{R}(x) %\arrow{d}{coupling}[swap]{\boldsymbol{v}} \\
		\arrow{d}{coupling} \\
		 & \boldsymbol{v}(\boldsymbol{R}(x)) 
	\end{tikzcd}
	\caption{Multi-class classification using a coupling method $\boldsymbol{v}$}
	\label{fig:coupling}
\end{figure}

@@

\emph{Bayes covariance} is a notion introduced in  [\cite{vsuch2016bayes}]. It refers to the behavior of a probabilistic multi-class method built in one-vs-one fashion, when the priors change. Any Bayes classifier, whether binary or multiclass changes its prediction when the priors change. Let us describe the process. Suppose we change priors from $\Pi = (P_1, \ldots, P_k)$ to $\Pi'= (P'_1, \ldots, P'_K)$. Then the prediction of Bayes classifiers changes from $\boldsymbol{p}^\Pi= (p_1, \ldots, p_K)$ to the vector 
\begin{align}
\boldsymbol{p}^{\Pi'} \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr). \label{eq:changePrior}
\end{align}

When we use a coupling method and the prior changes, we  can apply \eqref{eq:changePrior} either before or after coupling. If the results are identical, for all initial pairwise data $\boldsymbol{R}$, then we say that the coupling method is \emph{Bayes covariant}.

Let us make this explicit. Let $\boldsymbol{v}$ be a coupling method. Denote by  $\pi$ the change of priors $\pi:\Pi \rightarrow \Pi'$. If we have a  binary classifier predicting distribution $D \propto (p, p')$ on two classes $C_i$ and $C_j$, then we define new distribution $\pi_2^{ij}(D)$ by requiring
$$
\pi_2^{ij}(D) \propto \biggl(p \frac{P'_i}{P_i},p' \frac{P'_j}{P_j}\biggr).
$$
We denote by $\boldsymbol{R}^\pi$ the matrix with entries $\pi_2^{ij}(r_{ij})$. The value $\boldsymbol{v}(\boldsymbol{R}^\pi)$ represents effect of change of priors if we apply \eqref{eq:changePrior} \emph{before} coupling.

On the other hand we can apply \eqref{eq:changePrior} after coupling. Say we have a distribution on $K$ classes
\begin{align}
D \propto (p_1, \ldots, p_K)
\end{align}
then we define $D^\pi$ by requiring 
\begin{align}
D^\pi \propto \biggl(p_1 \frac{P'_1}{P_1}, \ldots, p_K \frac{P'_K}{P_K}\biggr).
\end{align}
We now say that a method if Bayes covariant if
\begin{align}
\boldsymbol{v}(\boldsymbol{R}^\pi)=	\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi,
\end{align}
for any pairwise predictions satisfying Assumption \ref{ass:1}.

%Let $\boldsymbol{R}^\pi$ be the matrix of pairwise predictions when we apply \eqref{eq:changePrior} to each binary classifier.
%For Bayes covariant method $\boldsymbol{v}$ the following diagram commutes.
%
%\begin{figure}[!h]
%\centering
%\begin{tikzcd}
%	\boldsymbol{R} \arrow[r, "\pi"] \arrow[d, "coupling"] & \boldsymbol{R}^\pi \arrow[d, "coupling"] \\
%	\boldsymbol{v}(\boldsymbol{R}) \arrow[r, "\pi"] & \boldsymbol{v}(\boldsymbol{R}^\pi) =\boldsymbol{v}(\boldsymbol{R})^\pi
%\end{tikzcd}
%\caption{Diagrammatical description of Bayes covariance for a coupling method $\boldsymbol{v}$}
%\label{fig:bc}
%\end{figure}

@@

\begin{prop}
Any arboreal method is Bayes covariant.
\end{prop}
\begin{proof}
	We can proceed by induction. The statement is clear when $K=2$. Suppose the proposition is true for an integers $<K$ and we are given a spanning tree $T$ on $K$ vertices. Any tree has a vertex of degree one. We may assume that in the case of $T$ it is the vertex corresponding to $C_1$ and the unique edge incident to $C_1$ is $C_1C_2$.

	 Denote by $\boldsymbol{t}$ the prediction corresponding to $T$ and by $\boldsymbol{t}'$ the prediction corresponding to the subgraph without the vertex $C_1$. Let us write 
	\begin{align*}
	\boldsymbol{t}(x) &= (t_1, t_2, \ldots, t_K) \\
	\boldsymbol{t'}(x) &= ( t'_2, \ldots, t'_K) 
\end{align*}
These vectors are related as follows:
\begin{align}
\begin{split}
\frac{t_1}{t_1+t_2} &= r_{12} \\
(t_2, \ldots, t_K) &\propto (t'_2, \ldots, t'_{K})
\end{split} \label{eq:ind1}
\end{align}

Now consider the effect of change of priors $\Pi = (P_1, \ldots, P_K)$ to $\Pi' = (P'_1, \ldots, P'_K)$. Applying \eqref{eq:changePrior} to the multi-class prediction yields a vector 
$$
\boldsymbol{t}^{\Pi\rightarrow \Pi'} \propto (t_1 \frac{P'_1}{P_1}, \ldots, t_K \frac{P'_K}{P_K}).
$$
Now consider what happens when we apply \eqref{eq:changePrior} before coupling. By induction assumption the method $\boldsymbol{t}'$ is Bayes covariant and thus applying coupling \emph{after} change of priors yields
$$
\boldsymbol{t}'^{\Pi\rightarrow \Pi'} \propto (t_2 \frac{P'_2}{P_2}, \ldots, t_K \frac{P'_K}{P_K})
$$

\end{proof}

@@

Suppose real numbers $\alpha_1, \ldots,\alpha_L$ are real numbers and $\boldsymbol{v}_1, \ldots, \boldsymbol{v}_L$ are probabibilistic multi-class classification methods satisfying Assumption \ref{ass:1}. We define their linear combination $\bigoplus_i a_i \boldsymbol{v}_i$ as the classification method that for any $x$ yields the probability distribution on $L$ classes that satisfies
\begin{align*}
 	\bigl(a_1 \boldsymbol{v}_1 \oplus \ldots \oplus a_L \boldsymbol{v}_L\bigr)(x) \propto \boldsymbol{v}_1^{a_1}(x) \odot \cdots \odot \boldsymbol{v}_L^{a_L}(x),
\end{align*}
where the symbol $\odot$ denotes componentwise multiplication. With this definition we can state the following result.

@@

\begin{proof}

\end{proof}
